---
title: "R 01"
output: html_notebook
---

```{r}
library(MASS)
data(Cars93)
attach(Cars93)
```

# Frequency tables, pie charts and bar charts

Frequency table

```{r}
table(Type)
freq.tb.Type <- table(Type)
freq.tb.Type
```

Relative frequency table

```{r}
n <- length(Type)
rel.freq.tb.Type <- freq.tb.Type/n
rel.freq.tb.Type

round(rel.freq.tb.Type, digits=2)
```

Pie chart

```{r}
pie(freq.tb.Type)
```

Bar chart

```{r}
barplot(freq.tb.Type)
barplot(rel.freq.tb.Type)
```

More than one variable:

```{r}
tb.Origin.Type <- table(Origin, Type)
tb.Origin.Type

tb.Origin.Type['USA', 'Large']

barplot(tb.Origin.Type)
barplot(tb.Origin.Type, beside=TRUE)
```

Add a legend

```{r}
barplot(tb.Origin.Type, beside=TRUE, legend=TRUE)
```

# Histograms

```{r}
Rear.seat.room.cm <- Rear.seat.room * 2.54
hist(Rear.seat.room.cm)
```

Making the histogram look nicer: arguments xlab, ylab, main and col

```{r}
hist(Rear.seat.room.cm, xlab='cm', ylab='', main='Rear Seat Room', col='orange')
```

It may be convenient to define graphical parameters (such as text options) of a plotting function as variables to be used in all instances of function.

```{r}
rsr     <- 'Rear Seat Room'
rsr.x   <- 'cm'
rsr.y   <- ''
rsr.col <- 'orange'
```

Function `par()` to split graphic panel and specify number of bins: argument breaks.

```{r}
par(mfrow=c(1,2))

hist(Rear.seat.room.cm, xlab=rsr.x, ylab=rsr.y,  main=rsr, col=rsr.col)
  
hist(Rear.seat.room.cm, breaks=30, xlab=rsr.x, ylab=rsr.y,  main=rsr, col=rsr.col)

par(mfrow=c(1,1))
```

Density vs frequency histogram: two equivalent ways to produce a density histogram.

```{r}
hist(Rear.seat.room.cm, freq=FALSE) 
hist(Rear.seat.room.cm, probability=TRUE) 
```

Specify the endpoints of the class intervals: function `range` and argument `na.rm`

```{r}
range(Rear.seat.room.cm)
range(Rear.seat.room.cm, na.rm=TRUE)
```

Compare the histogram under four conditions:\
1. equispaced vs non-equispaced bins\
2. frequency vs density histogram

```{r}
bins <- c(48, 65, 70, 75,  92)

par(mfrow=c(2,2))

hist(Rear.seat.room.cm,
     freq=FALSE,
     main='density/equispaced',
     xlab='',
     ylab='',
     col=rsr.col)
  
hist(Rear.seat.room.cm,
     freq=TRUE,
     main='frequency/equispaced',
     xlab='',
     ylab='',
     col=rsr.col)

hist(Rear.seat.room.cm,
     freq=FALSE,
     breaks=bins,
     main='density/non-equispaced',
     xlab='',
     ylab='',
     col=rsr.col)
  
hist(Rear.seat.room.cm,
     freq=TRUE,
     breaks=bins,
     main='frequency/non-equispaced',
     xlab='',
     ylab='',
     col=rsr.col)
 
par(mfrow=c(1,1))
```

Output of the function `hist()`:

```{r}
hist(Rear.seat.room.cm)

h <- hist(Rear.seat.room.cm)
h

is.list(h)

h$counts
h[[2]]

class(h)
```

The generic function `plot()` has a specific method for objects of class 'histogram':

```{r}
plot(h)
```

# Density Plot

```{r}
dens <- density(Rear.seat.room.cm, na.rm=TRUE) # kernel estimation of the density
plot(dens, main='Kernel Density') # plot the estimate
```

Compare density plot with histogram

```{r}
hist(Rear.seat.room.cm, freq=FALSE)
lines(dens, lwd=2)
```

Change the bandwidth

```{r}
dens0.5 <- density(Rear.seat.room.cm, adjust=0.5, na.rm=TRUE)

hist(Rear.seat.room.cm, freq=FALSE)
lines(dens0.5, col='red',  lwd=2)
lines(dens, col='blue', lwd=2)
```

Take a look at the object `dens`

```{r}
class(dens)

is.list(dens)
dens$x[1:10]
dens$y[1:10]
```

# Numerical Summaries

Arithmetic mean

```{r}
mean(Rear.seat.room.cm)
mean(Rear.seat.room.cm, na.rm=TRUE)
```

Removing NA's

```{r}
# using the function is.na()
is.na(Rear.seat.room.cm)
Rear.seat.room.cm[!is.na(Rear.seat.room.cm)]

# using the function na.omit()
na.omit(Rear.seat.room.cm)
Rear.seat.room.cm.cleaned <- na.omit(Rear.seat.room.cm)
```

Deviations from mean

```{r}
bar.x <- mean(Rear.seat.room.cm.cleaned)
d <- Rear.seat.room.cm.cleaned - bar.x

# Check that deviations from the mean are centered about zero
hist(d)
sum(d)
```

Median

```{r}
median(Rear.seat.room.cm.cleaned)
sort(Rear.seat.room.cm.cleaned)[46]
```

Arithmetic mean is highly sensitive to data points that are significantly different from the rest of the dataset, either much larger or much smaller.

```{r}
data.points <- 1:10
data.points
mean(data.points)
median(data.points)

data.points[10] <- 100
data.points
mean(data.points)
median(data.points)

data.points[10] <- 1000
data.points
mean(data.points)
median(data.points)
```

Variance and standard deviation

```{r}
x <- Rear.seat.room.cm.cleaned
s2 <- sum((x-mean(x))^2)/(length(x)-1)
var(x)

sd(x)
sqrt(s2)
```

Quantiles

```{r}
quantile(Rear.seat.room.cm.cleaned, 0.3)
quantile(Rear.seat.room.cm.cleaned, c(0.3, 0.6))
quantile(Rear.seat.room.cm.cleaned)

# Interquartile range
IQR(Rear.seat.room.cm.cleaned)

# Five-number summary (and mean)
summary(Rear.seat.room.cm.cleaned)
```

# Scatterplots and Boxplots

Boxplot

```{r}
boxplot(Rear.seat.room.cm.cleaned)
boxplot(Min.Price)
```

Side-by-side boxplots

```{r}
Rear.seat.USA     <- Rear.seat.room.cm.cleaned[Origin=='USA']
Rear.seat.non.USA <- Rear.seat.room.cm.cleaned[Origin=='non-USA']

boxplot(Rear.seat.USA, Rear.seat.non.USA)
```

Model formula syntax

```{r}
boxplot(Rear.seat.room.cm.cleaned ~ Origin[!is.na(Rear.seat.room.cm)])

boxplot(Rear.seat.room.cm.cleaned~Type[!is.na(Rear.seat.room.cm)])
boxplot(Length~Type)

boxplot(Min.Price~Origin)

detach(Cars93)
```

'Old faithful' data and scatterplots

```{r}
data('faithful')
attach(faithful)
```

EDA of waiting time

```{r}
summary(waiting)
par(mfrow=c(1, 2))
hist(waiting, col='orange', freq=FALSE, main='')
boxplot(waiting, col='lightgray')
par(mfrow=c(1,1))
```

EDA of eruption duration

```{r}
summary(eruptions)

par(mfrow=c(1, 2))
hist(eruptions, col='orange', prob=TRUE, main='')
boxplot(eruptions, col='lightgray')
par(mfrow=c(1,1))

summary(faithful)
```

Scatterplots: function `plot()`

```{r}
plot(eruptions, waiting)
plot(waiting ~ eruptions)
```

Function `plot()`, choice of symbol: argument pch

```{r}
plot(1:20)
plot(1:20, pch='A')
plot(1:20, pch=16)
plot(1:20, pch=1:20)
plot(1:20, pch=3)

# Function plot(), choice of symbol: argument col
plot(1:20, col='red', pch=16)
colors()
```

Final scatterplot

```{r}
lx <- 'Eruption time (min)'
ly <- 'Waiting time to next eruption (min)'
lmain <- 'faithful data: Eruptions of Old Faithful'

plot(eruptions, waiting, xlab=lx, ylab=ly, main=lmain, pch='+')

# LOESS=LOcally Estimated Scatterplot Smoothing
int.line <- loess.smooth(eruptions, waiting, span=2/3) # span=2/3 is the default value
lines(int.line,  col='blue', lty=1, lwd=2)

int.line # this is a list with the x and y values

# smoother curve: span=2
int.line.2 <- loess.smooth(eruptions, waiting, span=2) 
lines(int.line.2,  col='red', lty=2, lwd=2)

# rougher curve: span=1/5
int.line.0.20 <- loess.smooth(eruptions, waiting, span=1/5) 
lines(int.line.0.20,  col='violet', lty=4, lwd=4)

# Add a legend
legend(x=1.5, y=95, 
       legend=c('smoothing=1/5','smoothing=2/3', 'smoothing=2'), 
       col=c('violet', 'blue', 'red'),
       lty=c(4,2,1), 
       lwd=c(4,2,2)
       )

# Alternative way for legend position
# 'bottomright', 'bottom', 'bottomleft', 'left', 
# 'topleft', 'top', 'topright', 'right',  'center'
legend('topleft', y=NULL,
       legend=c('smoothing=1/5','smoothing=2/3', 'smoothing=2'), 
       col=c('violet', 'blue', 'red'),
       lty=c(4,2,1), 
       lwd=c(4,2,2)
       )

# size of the legend
legend('topleft', , y=NULL, cex=0.8,
       legend=c('smoothing=1/5','smoothing=2/3', 'smoothing=2'), 
       col=c('violet', 'blue', 'red'),
       lty=c(4,2,1), 
       lwd=c(4,2,2)
       )
```

`scatter.smooth()` function: `loess.smooth()`+`plot()`

```{r}
scatter.smooth(eruptions, waiting, pch='+')
```

old faithful data are time series

```{r}
n <- length(waiting)
time <- 1:n

plot(time, waiting)
```

First 100 time points and plot with 'line'

```{r}
plot(time[1:100], waiting[1:100], type='l')


waiting.next <- waiting[-1]
waiting.previous <- waiting[-n]

plot(waiting.previous, waiting.next, pch='+')

abline(h=70, lty=2, col='red')
abline(v=70, lty=2, col='red')

detach(faithful)
```

# 'Animals' data and transformations

```{r}
library(MASS)
data(Animals)
attach(Animals)
```

```{r}
summary(body)
hist(body, prob=TRUE)
hist(body, breaks=20, prob=TRUE, col=7)
```

Logarithmic transformation

```{r}
lbody <- log10(body)
summary(lbody)
hist(lbody, prob=TRUE, col=7)

# Inverse transformation
10^mean(lbody)
summary(body)
```

Geometric mean

```{r}
n <- length(body)
prod(body)^(1/n)
```

Scatterplot

```{r}
plot(body, brain, pch='+')

lbrain <- log10(brain)
plot(lbody, lbrain, pch=16)
```
