---
title: "R 02"
output: html_notebook
---

# Bernoulli Random Variable

```{r}
# Bernoulli trial with probability of success equal to 0.7
urn <- c(rep('SUCCESS', 7), rep('FAILURE', 3))
urn

# Simulate random extraction from a urn
sample(urn, 1)

# n independent replications of a Bernoulli trial
n <- 4
sample(urn, size=n, replace = TRUE)

# Random seed: state of the random number generator in R
set.seed(321)
sample(urn, size=n, replace = TRUE)
```

Bernoulli random variable with prob=0.7

```{r}
urn <- c(rep(1, 7), rep(0, 3))
urn

# one observation 
sample(urn, 1) 

# n independent realizations
n <- 10
sample(urn, size=n, replace = TRUE)
```

# Binomial Distribution

Generate one realization from a Binomial random variable n=10 and prob=0.7

```{r}
n <- 10
n.iid.bernoulli <- sample(urn, size=n, replace = TRUE)
n.iid.bernoulli
sum(n.iid.bernoulli)
```

This can be done with the rbinom() function

```{r}
# one random value
rbinom(1, size=10, prob=0.7)

# four random values
rbinom(4, size=10, prob=0.7)
```

Binomial pdf and cdf

```{r}
p <- 0.7
n <- 10

# p(X=3)
dbinom(3, n, p)

# probability mass function 
x <- 0:n
pdf <- dbinom(x, n, p)
pdf

sum(pdf) # check that the sum is 1
```

a plot of the probability mass function

```{r}
plot(x, pdf, xlab = 'x', ylab = 'p(x)', type = 'h')

# Function expression() to improve the quality of axis label (xlab and ylab)
plot(x, pdf, xlab = expression(x), ylab = expression(p(x)), type = 'h')
```

Rules for using the expression() function

```{r}
# Run in terminal
# demo(plotmath)
```

```{r}
# P(X<=3)
pbinom(3, n, p)

# cumulative distribution function
x <- 0:n
cdf <- pbinom(x, n, p)
cdf

```

Plot cumulative distribution function

```{r}
# basic step function
n <- 10
x <- 0:n
cdf <- pbinom(x, n, p)
out.step <- stepfun(x, c(0, cdf))
out.step

plot(out.step)

# nicer output with more graphical parameters

plot(out.step,
     pch=20,
     vertical=FALSE,
     xlab=expression(x),
     ylab=expression(F(x)==P(X<=x)),
     main='')
```

# EXAMPLE: AA Airlines

User defined functions

```{r}
my.square <- function(a){
  b <- a*a
  return(b)
}

my.square

my.square(2)
x <- my.square(2)
x
```

'for' loops

```{r}
# example 1
for(i in 1:5) print(i)

# example 2
x <- rbinom(5, size=10,  p=0.5)
for (k in x) print(k)

# example 3
superheroes <- c('superman', 'batman', 'spiderman')
for (name in superheroes ) print(name)
```

Apply a function to a vector

```{r}
a <- c(1, 3, 5, 7)

# many (but not all) functions  work also with vectors 
x <- my.square(a)
x

# in alternative use a cycle
x <- c()
for(i in a) x <- c(x, my.square(i))
x

# sapply() function: the same ascycle for above but more efficient
x <- sapply(a, FUN = my.square)
x
```

Plot the binomial mass functions for different values of probability parameter

```{r}
# function that produces a plot of binomial(10, p)
binom.plot <- function(p) {
  plot(0:10, dbinom(0:10,10,p), ylim=c(0,0.5), type='h')
  Sys.sleep(0.1)
  }

# plot for p=0.2
binom.plot(0.2)

# plot for p=0.8
binom.plot(0.8)

# 100 equispaced probabilities between 0 and 1
p.vec <- seq(0, 1, length=100)

# Run in terminal
# iterate the plot over the values in p.vec 
# for (p in p.vec) binom.plot(p)

# the same can be done with teh sapply() function
# ignore <- sapply(p.vec, binom.plot)
```

Another example: behavior of binomial as n increases function for binomial(n, 0.5)

```{r}
# Run in terminal
# binom.plot <- function(n) {
#   plot(0:n, dbinom(0:n,n,0.5), type='h')
#   Sys.sleep(0.1)
# }
# 
# ignore <- sapply(1:100, binom.plot)
```

# Poisson Distribution

```{r}
yl <- expression(p[Y](y))

par(mfrow=c(2,2))

y <- 0:7
plot (y, dpois(y,1), xlab='y', ylab=yl, main=expression(lambda==1), type='h')

y <- 0:10
plot (y, dpois(y,4), xlab='y', ylab=yl, main=expression(lambda==4), type='h')

y <- 6:26
plot (y, dpois(y,16), xlab='y', ylab=yl,main=expression(lambda==16), type='h')

y <- 44:84
plot (y, dpois(y,64), xlab='y', ylab=yl,main=expression(lambda==64), type='h')

par(mfrow=c(1,1))
```
