---
title: "R 03"
output: html_notebook
---

# Uniform Distribution

```{r}
# density function of U(0, 1)
dunif(0.2, min=0, max=1)
dunif(0.2) # 0 and 1 are the default values

# cumulative distribution function of U(0, 1)
punif(0.2)

# uniformly distributed random values
runif(3)
runif(3, -3, 4)
```

# Exponential Distribution

```{r}
f <- function(x, lambda=1) lambda*exp(-lambda*x)

f(0.5)
f(0.5, lambda=1/3)

# or equivalently

dexp(0.5)
dexp(0.5, rate=1/3) # rate=lambda
```

Plot Probability Density Function

```{r}
# the curve function 
curve(f, from=0, to=5) # note that f needs to be a function of 'x'

curve(f(x, lambda=1), 0, 5) # to include additional arguments 
curve(f(x, lambda=1/3), 0, 5)

# exploit the build-in function for pdf
curve(dexp, 0, 5)
curve(dexp(x, rate=1/3), 0, 5)

# plot the cdf
curve(pexp, 0, 5)
curve(pexp(x, rate=1/3), 0, 5)
```

Comparison of exponential densities

```{r}
x <- seq(0, 2, length=40)
theta <- c(2, 1, .2, .1) # mean of distribution
y <- matrix(NA, 40, 4)
for (i in 1:4) {
  y[,i] <- dexp(x, 1/theta[i]) # parameter is the rate=lambda
}

# legend with the expected value 'theta'
matplot(x, y, type='l', xlab='x', ylab='p(x)', lty=1:4, col=1)
legend('topright', paste('theta =', theta), lty=1:4)

# legend with the rate 'lambda'
matplot(x, y, type='l', xlab='x', ylab='p(x)', lty=1:4, col=1)
legend('topright', paste('lambda =', 1/theta), lty=1:4)
```

# Normal Distribution

Distribution for different values of mu and sigma

```{r}
# some pdf's
curve(dnorm,-8,8)
curve(dnorm(x, mean=0, sd=2), add=TRUE, col=2)
curve(dnorm(x, mean=0, sd=4), add=TRUE, col=3)

curve(dnorm, -8, 8)
curve(dnorm(x, -2, 1), add=TRUE, col=2)
curve(dnorm(x,  2, 2), add=TRUE, col=3)

# and corresponding cdf's
curve(pnorm,-8,8)
curve(pnorm(x,0,2),add=TRUE, col=2)
curve(pnorm(x,0,4),add=TRUE, col=3)

# compute probabilities of the normal distribution 
# N(0, 1), Pr(X<0)
pnorm(0)

# N(0,1), Pr(-1<X<2)
pnorm(2)-pnorm(-1)

# N(2, 5), Pr(1<X<7)
pnorm(7, mean = 2, sd=sqrt(5))-pnorm(1, mean = 2, sd=sqrt(5))
```

# Quantiles and Cumulative Distribution Function (CDF)

Uniform distribution

```{r}
qunif(0.6, min=-0.5, max=1.5)

pr <- 0.6
q  <- qunif(pr, min=-0.5, max=1.5)
q
```

Represent the quantile on the probability DENSITY function

```{r}
curve(dunif(x, min=-0.5, max=1.5), -0.5, 1.5, xlim=c(-1, 2), ylim=c(0, 0.7), ylab='f(x)')
lines(c(-0.5, -0.5), c(0, 0.5), lty=3)
lines(c(1.5, 1.5), c(0, 0.5), lty=3)
abline(h=0)
polygon(c(-0.5, -0.5, q, q), c(0, 0.5, 0.5, 0), col='gray')
```

Represent the quantile on the CUMULATIVE distribution function

```{r}
pr <- 0.6
q <- qunif(pr, -0.5, 1.5)
curve(punif(x, min=-0.5, max=1.5), -1, 2, ylab='F(x)', ylim=c(0,1.2))
abline(h=pr, v=q, lty=3, col='red')
```

Exponential distribution

```{r}
# Recall that for lambda=1 we have expected.value=standard.deviation=1
pr <- 0.6
q <- qexp(pr, rate=1)
q

# Quantile on the density function
par(mfrow=c(1, 2))

curve(dexp, 0, 4, ylab='f(x)')
x <- seq(0, q, length=1000)
y <- dexp(x)
polygon(c(0, x, q), c(0, y, 0), col='gray')

# Quantile on the cumulative distribution function
curve(pexp, 0, 4, ylim=c(0, 1.1), ylab='F(x)')
lines(x=c(0, q), y=c(pr, pr), lty=2, col='blue')
lines(x=c(q, q), y=c(0, pr), lty=2, col='blue')
par(mfrow=c(1, 1))
```

```{r}
# we consider mu=1 and sigma=1 so that we have same mean and std of exp(1) distribution 
pr <- 0.6
q <- qnorm(pr, mean=1, sd=1)
q
# Quantile on the density function
par(mfrow=c(1, 2))

curve(dnorm(x, mean=1, sd=1), -2, 4, ylab='f(x)')
x <- seq(-2, q, length=1000)
y <- dnorm(x, mean=1, sd=1)
polygon(c(-2, x, q), c(0, y, 0), col='gray')

# Quantile on the cumulative distribution function
curve(pnorm(x, mean=1, sd=1), -2, 3, ylim=c(0, 1.1), ylab='F(x)')
lines(x=c(-2, q), y=c(pr, pr), lty=2, col='blue')
lines(x=c(q, q), y=c(0, pr), lty=2, col='blue')
par(mfrow=c(1, 1))
```

Compare the quantile of normal distribution with that of exponential distribution

```{r}
pr <- 0.6

curve(pnorm(x, mean=1, sd=1), -2, 3, ylim=c(0, 1.1), ylab='F(x)')
curve(pexp, -2, 3, ylim=c(0, 1.1), ylab='F(x)', add=TRUE, col='blue')

abline(h=pr, lty=3, col='red', lwd=2)
points(c(qexp(pr), qnorm(pr, mean=1, sd=1)), c(pr, pr), pch=16)
```

empirical cdf

```{r}
set.seed(123)
x <- rexp(4)
sort(x)

plot(ecdf(x), xlim=c(0,1.75), ylim=c(0, 1.1), axes=FALSE)
box()
axis(2, at=c(0, 0.25, 0.5, 0.75, 1))
axis(1, at=round(x, 3))
```

Exponential distribution: comparison between ecdf and 'exact' cdf

```{r}
set.seed(123)
x <- rexp(30)
plot(ecdf(x), xlim=c(0,4))
curve(pexp, add=TRUE, col='blue')
```

normal distribution: comparison between ecdf and 'exact' cdf

```{r}
set.seed(123)
x <- rnorm(30, mean=1, sd=1)
plot(ecdf(x), xlim=c(-3,3))
curve(pnorm(x, mean=1, sd=1), add=TRUE, col='blue')
```

comparison between ecdf from exponential and cdf from standard normal

```{r}
set.seed(123)
x <- rexp(30)
plot(ecdf(x), xlim=c(-3,3))
curve(pnorm(x, mean=1, sd=1), add=TRUE, col='blue')
```

Empirical cdf

```{r}
set.seed(123)
x <- rexp(30)
qqnorm(x)
qqline(x)
```

Comparison with different cases

```{r}
# right skewed

set.seed(123)
x <- rexp(50)

par(mfrow=c(1,2))
hist(x, freq=FALSE, col='cyan')
qqnorm(x)
qqline(x)
par(mfrow=c(1,1))

# left skewed

set.seed(123)
x <- -rexp(50)

par(mfrow=c(1,2))
hist(x, freq=FALSE, col='cyan')
qqnorm(x)
qqline(x)
par(mfrow=c(1,1))


# light tails

set.seed(123)
x <- runif(30)


par(mfrow=c(1,2))
hist(x, freq=FALSE, col='cyan')
qqnorm(x)
qqline(x)
par(mfrow=c(1,1))

# heavy tails

set.seed(123)
x <- rt(50, 3)

par(mfrow=c(1,2))
hist(x, freq=FALSE, col='cyan')
qqnorm(x)
qqline(x)
par(mfrow=c(1,1))
```

# Data animals and z.scores

```{r}
library(MASS)
data(Animals)
attach(Animals)
```

```{r}
l.body <- log10(body)


par(mfrow=c(1, 2))
qqnorm(body)
qqline(body)
qqnorm(l.body)
qqline(l.body)
par(mfrow=c(1,1))
```

# Faithful gayser data and bimodality

```{r}
data(faithful)
attach(faithful)
```

```{r}
par(mfrow=c(1,2))
hist(waiting, freq=FALSE, col='cyan')
qqnorm(waiting)
qqline(waiting)
par(mfrow=c(1,1))
```

# Probability distributions related to normal one

Chi-squared distribution

```{r}
# density function with 5 degrees of freedom
curve(dchisq(x, df=5), xlim=c(0, 20))
```

Student's t distribution

```{r}
# 5 degrees of freedom
curve(dt(x, 5),  xlim=c(-5, 5), ylab='pdf', xlab='', col='blue', lwd=1.5)
```

Comparison with the normal distribution

```{r}
curve(dnorm, xlim=c(-5, 5), ylab='pdf', xlab='', col='red', lwd=1.5)
curve(dt(x, 1),  add=TRUE, lty=2, col='blue', lwd=1.5)
curve(dt(x, 4),  add=TRUE, lty=3, col='black', lwd=1.5)
curve(dt(x, 16), add=TRUE, lty=4, col='black', lwd=1.5)


leg.label <- c('t df = 1', 't df = 4', 't df = 16', 'Normal' )
leg.lty   <- c(2, 3, 4, 1)
leg.col <- c('blue', 'black', 'black',  'red')
legend (x='topleft', y=NULL, legend=leg.label, lty=leg.lty, col=leg.col, lwd=1.5)
```

Compare boxplots

```{r}
x <- rnorm(1000)
y <- rt(1000, 5)

boxplot(x, y, names=c('normal dist.', 't dist. with 5df'), col='lightgray')
```

Density function with 5 degrees of freedom

```{r}
curve(df(x, df1=5, df2=7), xlim=c(0, 5))
```

# Linear combination of rv's and central limit theorem

Mean of a sample from a normal random variable for different sample sizes

```{r}
# This function considers a random variable X~N(5,1) and plots its density function together with the density function of sample mean of n i.i.d. observations from X 
samplingdist <- function(n) {
  curve(dnorm(x,5,1), xlim=c(2,8), ylim=c(0,2), col=2, lwd=1.5, ylab='')
  curve(dnorm(x,5,1/sqrt(n)), xlim=c(2,8), ylim=c(0,2), lwd=1.5,add=TRUE)
  Sys.sleep(0.2)}

# apply samplingdist() for a single value of n
samplingdist(5)

# apply the function samplingdist() for n from 1 to 20 
# Run in terminal
# ignore <- sapply(1:20, samplingdist)
```

Central Limit Theorem (CLT)

```{r}
# Empirical check of the CLT with exponential distribution X~Exp (1) so that E(X)=var(X)=1
# We first consider a sample from Exp(1) and compare histogram of sample data both with true density function, that is of Exp(1) and with that of a normal distribution with same mean and variance N(1,1). 
y <- rexp(10000)
hist(y, prob=TRUE, main='Exp(1) vs N(0,1)', xlab='', ylab='', xlim=c(-2, 7), ylim=c(0, 1.1))
abline(v=0)
curve(dexp, add=TRUE, col='red', lwd=2, from=0.001, to=7)
curve(dnorm(x, mean=1, sd=1), add=TRUE, lwd=2, col='blue')
```

```{r}
# We now consider Y that is obtained from arithmetic mean of n i.i.d. Exp(1) random variables so that E(Y)=1, var(Y)=1/n and compare its denisty with that of N(1; 1/n) (note that sd=1/sqrt(n)) increasing values of n are considered set the value of n
n <- 2

# sample n i.i.d. values from X~Exp (1) and compute the mean
y <- mean(rexp(n))

# repeat this operation 10000 times and plot the histogram
y <- c()
for(i in 1:10000) y <- c(y, mean(rexp(n)))
hist(y, prob=TRUE, main=paste('n = ', n), xlab='', ylab='')

# add the density of the Normal approximation 
curve(dnorm(x, mean=1, sd=1/sqrt(n)), add=TRUE, lwd=2, col='blue')
```

```{r}
# implement this procedure in a function 
CLT.check <- function(n){
  mu.y <- 1
  sd.y <- 1/sqrt(n)
  y <- c()
  for(i in 1:10000) y <- c(y, mean(rexp(n)))
  hist(y, breaks=40, prob=TRUE, main=paste('n =', n), xlab='', ylab='', xlim=c(1-3*sd.y, 1+4*sd.y))
  curve(dnorm(x, mean=mu.y, sd=sd.y), add=TRUE, col='blue', lwd=2)
}


# compare what happens for increasing values of n
par(mfrow=c(2,3))
set.seed(123)
for (n in c(1, 2, 4, 20, 50, 100)) CLT.check(n)
par(mfrow=c(1,1))
```

# Central Limit Theorem: Exercises (from Exercises_01)

EXERCISE n.1

```{r}
# point a)
mu    <- 12.01
sigma <- 0.2

# mean of n=144 bottles
n <- 144
var.x.bar <- sigma^2/n
sd.x.bar  <- sqrt(var.x.bar)

# note that 144=12^2
sd.x.bar <- sigma/12
pnorm(12, mean=mu, sd=sd.x.bar)

# point b) 
mu <- 12.3
pnorm(12, mean=mu, sd=sd.x.bar)
```

EXERCISE n.2

```{r}
mu    <- 0.08
sigma <- 0.01

n <- 250

mu.y  <- n*mu
var.y <- n*sigma^2
sd.y  <- sqrt(var.y)

# a)
1-pnorm(20.2, mean=mu.y, sd=sd.y)

# b) 
qnorm(0.1, mean=mu.y, sd=sd.y)
qnorm(0.9, mean=mu.y, sd=sd.y)

# c) There is not enough information: the distribution of page thickness in unknown
```

EXERCISE n.3

```{r}
x <- 0:4
pr <- c(0.1, 0.3, 0.3, 0.2, 0.1)

# expected value of X
mu <- sum(x*pr)
mu

# variance of X
sigma2 <- sum((x-mu)^2*pr)
sigma2

# mean of 100 days
n <- 100
var.x.bar <- sigma2/n
sd.x.bar  <- sqrt(var.x.bar)
1-pnorm(2, mean = mu, sd = sd.x.bar)
```

EXERCISE n.9

```{r}
q.95 <- qnorm(0.95)
q.95

n <- (q.95*0.5/0.1)^2
n
```

EXERCISE n.10

```{r}
n <- 100
pr <- 0.3

# exact probability form Binomial distribution
1 - pbinom(35, n, pr)

# approximate probability using the CLT
mu.y <- n*pr
sigma2.y <- n*pr*(1-pr) 

1 - pnorm(35, mean=mu.y, sd=sqrt(sigma2.y))


# compare the exact Binomial pdf with approximating Normal pdf
x <- 10:50
pr.x <- dbinom(x, n, pr)
plot(x, pr.x, type='h', xlab='', ylab='', main='')
curve(dnorm(x,  mean=mu.y, sd=sqrt(sigma2.y)), add=TRUE, lwd=2, lty=2, col='red')

# note that a better approximation could be obtained by applying a 'continuity correction' that is done by replacing 35 with 35.5
1 - pnorm(35.5, mean=mu.y, sd=sqrt(sigma2.y))
```

# Loading data into R

```{r}
setwd('/home/hieu/BDMA/SL/')

temp.data <- read.table('normtemp-data/normtemp.txt', head=TRUE)
attach(temp.data)
temp.C <- (temperature-32)*5/9

par(mfrow=c(1, 2))
qqnorm(temp.C, main='body temperature')
qqline(temp.C)


par(mfrow=c(1,1))
```

Missing values

```{r}
# default na.strings = 'NA'
mydata <- read.table('normtemp-data/normtemp-with-NA.txt', head=TRUE)

mydata$temperature[1:10]
is.vector(mydata$temperature)
is.character(mydata$temperature)
is.numeric(mydata$temperature)

# set na.strings='*'
mydata <- read.table('normtemp-data/normtemp-with-NA.txt', head=TRUE, na.strings = '*')

mydata$temperature[1:10]
is.vector(mydata$temperature)
is.character(mydata$temperature)
is.numeric(mydata$temperature)
```

Factors (i.e. data type for categorical variables)

```{r}
is.vector(mydata$gender)
is.character(mydata$gender)
is.factor(mydata$gender)

# define a factor
gender.f <- as.factor(mydata$gender)
is.factor(gender.f)
```

```{r}
# different method  (i.e. behavior) for printing character vectors and factors
mydata$gender[1:10]
gender.f[1:10]

# different method (i.e. behavior) for summary() of character vectors and factors
summary(mydata$gender)
summary(gender.f)
```

```{r}
# modify directly the attributes of the data.frame
mydata$gender <- as.factor(mydata$gender)
```

```{r}
# ordered factor 
mydata <- read.table('normtemp-data/normtemp-with-ordinal-var.txt', head=TRUE, comment.char = '#')

is.character(mydata$age)
is.factor(mydata$age)

# convert into a factor (categorical variable)
age.f <- factor(mydata$age)
is.vector(age.f)
is.factor(age.f)
```

```{r}
# check whether it is recognized as ordered
is.ordered(age.f)

# the levels are printed in arbitrary order 
age.f[1:10]

# convert into an ordered factor
age.fo <- factor(mydata$age, ordered=TRUE, levels=c('<30', '[30, 50)', '[50, 70)', '>=70'))

# the print() function now highlights the order of levels
age.fo[1:10]
```

```{r}
is.ordered(age.fo)
summary(age.f)
summary(age.fo)

# modify directly the data.frame
mydata$age <- factor(mydata$age, ordered=TRUE, levels=  c('<30', '[30, 50)', '[50, 70)', '>=70' ))
```
