---
title: "R 07"
output: html_notebook
---

# Comparison of estimators based on variance/standard errors

Following the body-temperature example, we assume the population normally distributed

```{r}
mu <- 36.75
sigma <- 0.45

# we consider three different estimators of mu:
#
# mu.hat.1 = sample mean
# mu.hat.2 = sample median
# mu.hat.3 = midpoint of the range of the sample
#
# hence simulate the sampling distributions of these 
# three estimators based on 10^4 samples with n=16

samplesize <- 16

mu.hat.1 <- c()
mu.hat.2 <- c()
mu.hat.3 <- c()

for (i in 1:10000) {
  x <- rnorm(samplesize, mean=mu, sd=sigma)
  mu.hat.1 <- c(mu.hat.1, mean(x))
  mu.hat.2 <- c(mu.hat.2, median(x))
  mu.hat.3 <- c(mu.hat.3, mean(range(x)))
}
```

```{r}
par(mfrow=c(2,2)) 
hist(mu.hat.1, main=expression(hat(mu)[1]~(mean~of~values)), freq=F, col="cyan", ylab="")
hist(mu.hat.2, main=expression(hat(mu)[2]~(median~of~values)), freq=F, col="cyan", ylab="")
hist(mu.hat.3, main=expression(hat(mu)[3]~(mean~of~range)), freq=F, col="cyan", ylab="")
d1 <- density(mu.hat.1)
d2 <- density(mu.hat.2)
d3 <- density(mu.hat.3)
plot(d1, lwd=2, main="", xlim=range(mu.hat.3), xlab="", ylab="")
abline(v=mu, lty=2)
lines(d2, lwd=2, col="blue")
lines(d3, lwd=2, col="red")
par(mfrow=c(1,1))
```

```{r}
# comparison of empirical distributions 
plot(d1, lwd=2, main="", xlim=range(mu.hat.3), xlab="", ylab="")
legend("topleft", lty=c(1,1,1), col=c("black", "blue", "red"), cex=0.75, legend=c("mean", "median", "mid-range"))
abline(v=mu, lty=2)
lines(d2, lwd=2, col="blue")
lines(d3, lwd=2, col="red")
```

```{r}
# compute the mean and the standard deviations 
# (i.e. the standard errors) of the estimators
mean(mu.hat.1)
sd(mu.hat.1)

mean(mu.hat.2)
sd(mu.hat.2)

mean(mu.hat.3)
sd(mu.hat.3)
```

# MUON DECAY EXAMPLE

```{r}
# numerical computation of MLE
# and of its asymptotic distribution
muon.data <- c(0.41040018,  0.91061564, -0.61106896,  0.39736684,  0.37997637, 0.34565436,
               0.01906680, -0.28765977, -0.33169289,  0.99989810, -0.35203164, 0.10360470,
               0.30573300,  0.75283842, -0.33736278, -0.91455101, -0.76222116, 0.27150040,
               -0.01257456,  0.68492778, -0.72343908,  0.45530570,  0.86249107, 0.52578673,
               0.14145264,  0.76645754, -0.65536275,  0.12497668,  0.74971197, 0.53839119)

# function for density of muon decay
dmuon <- function(x, alpha){
  d <- (1+alpha*x)/2
  return(d)
}

# compute dmuon() for some values of x
x <- muon.data[1]
x
dmuon(x, alpha=0.7)

# this function works also if x is a vector
x <- muon.data[1:4]
dmuon(x, alpha=0.7)
```

```{r}
# plot the density for some alpha values
a <- 0.7 # positive
a <- 0   # zero
a <- -0.7 # negative

curve(dmuon(x , alpha=a), from=-1, to=1, ylab=expression(p(x)), main='Density for muon decay')
```

```{r}
# function for log-likelihood
lmuon <- function(alpha, data){
  l <- sum(log(1+alpha*data))-length(data)*log(2)
  return(l)
}

# compute for some value of alpha
lmuon(-0.4, data=muon.data)

# this function doesn't work as expected  if one tries to use  it with a vector of values or plot a curve. 
# alpha <- c(-0.4, 0.2, 0.7)
# lmuon(alpha, data=muon.data)
# 
# curve(lmuon(x, data=muon.data), -1, 1)

# we need to "vectorize" the function
lmuon.v <- function(alpha, data){
  l <- c()
  for (a in alpha) l <- c(l, lmuon(a, data))
  return(l)
}

lmuon(-0.4, data=muon.data)
alpha <- c(-0.4, 0.2, 0.7)
lmuon.v(alpha, data=muon.data)

xl <- expression(alpha)
yl <- expression(l(alpha))
mt <- "log-likelihood for muon decay"

curve(lmuon.v(x, data=muon.data), from=-1, to=1, xlab=xl, ylab=yl, main=mt)

# "Vectorize" function
lmuon.v <- Vectorize(FUN=lmuon, vectorize.args = "alpha")
lmuon.v

lmuon.v(alpha, data=muon.data)
curve(lmuon.v(x, data=muon.data), from=-1, to=1, xlab=xl, ylab=yl, main=mt)
```

```{r}
# score function
smuon <- function(alpha, data){
  l <- sum(data/(1+alpha*data))
  return(l)
}

smuon.v <- Vectorize(FUN=smuon, vectorize.args = "alpha")
curve(smuon.v(x, data=muon.data), -0.8, 1, xlab=xl)
abline(h=0, lty=2)
```

# Built-in functions for root/optimization

```{r}
# The function uniroot() searches the interval from
# lower to upper for a root (i.e., zero)
uniroot(f=smuon, data=muon.data, lower=-1, upper=1)

uniroot(f=smuon, data=muon.data, interval=c(-1,1))


# optimization functions that by default perform minimization
minus.lmuon <- function(alpha, data) -lmuon(alpha, data)

nlminb(start=0.6, objective=minus.lmuon, data=muon.data, lower=-1, upper=1)

optim(0.6, fn=minus.lmuon, data=muon.data, lower=-1, upper=1)
optim(0.6, fn=minus.lmuon, method="Brent",data=muon.data, lower=-1, upper=1)
```

# Muon decay - asymptotic distribution MLE

```{r}
# recall the log-likelihood function
lmuon <- function(alpha, data){
  l <- sum(log(1+alpha*data))-length(data)*log(2)
  return(l)
}

minus.lmuon <- function(alpha, data) -lmuon(alpha, data)

# compute the MLE alpha.hat of alpha
alpha.hat <- nlminb(0.6, minus.lmuon, data=muon.data, lower=-1, upper=1)$par

# compute the observed information at alpha.hat
j <- sum(muon.data^2/(1+alpha.hat*muon.data)^2)
j

# asymptotic approximation of the standard error
se <- sqrt(1/j)
se

# approximate 95% confidence interval
alpha.hat +c(-1, 1)*1.96*se

# is the distribution uniform over the interval (-1;1)?
#
# H_0: alpha=0
# H_1: alpha!=0
t.obs <- (alpha.hat-0)/se
p.value <- 2*pnorm(-abs(t.obs))
p.value
```

```{r}
# numerical derivative
library(numDeriv)

H <- hessian(func = lmuon, x = alpha.hat, data = muon.data)
j <- -H[1,1]
sqrt(1/j)
```

# Example: survival time for leukemia

```{r}
# data 
x <- c(56,65,17,7,16,22,3,4,2,3,8,4,3,30,4,43)
```

Computation of MLEs as done on the slides

```{r}
score <- function(beta, data)
{
  n <- length(data)
  n/beta +sum(log(data))-n*(sum(x^beta*log(data))/(sum(data^beta)))
}

hat.beta <- uniroot(score, data=x, interval=c(0.1,100))$root
hat.beta
hat.alpha <- (mean(x^hat.beta))^(1/hat.beta)
hat.alpha
```

(alternative) computation of MLEs using only numerical procedures

```{r}
# density of Weibull distribution (as given in the slide)
dW <- function(x, alpha, beta) beta*alpha^-1*(x/alpha)^(beta-1)*exp(-(x/alpha)^beta)
dW(1, alpha=1, beta=2)

# equivalent build-in function

# shape = beta
# scale = alpha

dweibull(1, scale=1, shape=2)


# log-likelihood function
#
# Note that:
# 1.  the parameters are given as a single vector and 
#     as first argument, this is a requirement of the function 
#     nlminb() used later for optimization
#
# 2. we exploit the function dW() so as to avoid the 
#    explicit computation (and implementation) of the
#    log-likelihood function

loglik.W <- function(par, data){#par=c(alpha, beta)
  alpha <- par[1]
  beta  <- par[2]
  l <- sum(log(dW(data, alpha, beta))) 
  return(l)
}

# change the sign of the log-likelihood function 
# (required because the function nlminb() is a minimizer)
#
minus.loglik.W <- function(par, data) -loglik.W(par, data)


# Numerical computation of the Maximum Likelihood Estimates 
# of alpha and beta
#
opt.result <- nlminb(start = c(1,1), objective = minus.loglik.W, data=x)
opt.result$par

# store the estimate as scalars
alpha.hat <- opt.result$par[1]
beta.hat  <- opt.result$par[2]
alpha.hat
beta.hat
```

```{r}
# Compute the observed information matrix
# with a function for numerical derivatives

library(numDeriv)

# Hessian Matrix
H <- hessian(func = loglik.W, x = c(alpha.hat, beta.hat), data=x)
H

# observed Fisher information  matrix
J <- -H
J

# asymptotic variance and covariance matrix

asy.var <- solve(J)
asy.var

# (approximate) standard error of parameters

se.alpha <- sqrt(asy.var[1,1])
se.beta  <- sqrt(asy.var[2,2])

# 95% (approximate) confidence intervals

alpha.hat+c(-1,1)*qnorm(0.975)*se.alpha
beta.hat +c(-1,1)*qnorm(0.975)*se.beta

# memoryless property?
#
# H_0: beta=1
# H_1: beta!=1

t.obs <- (beta.hat-1)/se.beta
t.obs

p.value <- 2*pnorm(-abs(t.obs))
p.value
```
