---
title: "Lab 1: Linear regression and Nonlinear regression"
output: html_notebook
---

```{r}
Sys.setlocale('LC_ALL', 'C')
library(fpp2)
library(readxl) # read excel
library(lmtest) # linear model test
library(DIMORA) # BASS model
setwd('/home/hieu/BDMA/TSA/')
```

# Some plots for time series

```{r}
plot(a10, ylab='million dollars', xlab='Year', main='Antidiabetic drugs')
```

```{r}
tsdisplay(a10) ##a general view of the data
```

```{r}
seasonplot(a10, ylab='million dollars', xlab='Year', main='Seasonal plot: Antidiabetic drugs', year.labels=T, year.labels.left=T, col=1:20, pch=19)
```

# Linear regression for time series

Facebook example

```{r}
# read the data
facebook<- read_excel('facebook.xlsx')
str(facebook)
# create a variable 'time'
tt<- 1:NROW(facebook)

# create the variable 'fb'
fb <- facebook$fb

# make a plot
plot(tt, fb, xlab='Time', ylab='Facebook users')
```

```{r}
# acf of variable 'fb'
acf(fb)

# fit a linear regression model 
fit1 <- lm(fb~ tt)
summary(fit1)
anova(fit1)

# plot of the model
plot(tt, fb, xlab='Time', ylab='Facebook users')
abline(fit1, col=3)
```

```{r}
# check the residuals? are they autocorrelated? Test of DW
dwtest(fit1)

# check the residuals
resfit1<- residuals(fit1)
plot(resfit1,xlab='Time', ylab='residuals' )
```

```{r}
# let us do the same with a linear model for time series, so we transform the data into a 'ts' object
fb.ts <- ts(fb, frequency = 4)
ts.plot(fb.ts, type='o')

# we fit a linear model with the tslm function
fitts<- tslm(fb.ts~trend)

# obviously it gives the same results of the first model
summary(fitts)

dwtest(fitts)
```

Linear regression for iMac

```{r}
apple<- read_excel('apple.xlsx')
str(apple)
imac <- apple$iMac

# data visualization
plot(imac,type='l', xlab='quarter', ylab='iMac sales')
```

```{r}
# variable tt for a linear model 
tt<- 1:NROW(apple)

# linear model
fit2 <- lm(imac~tt)
summary(fit2)

plot(imac,type='l', xlab='quarter', ylab='iMac sales')
abline(fit2, col=3)
```

```{r}
dwtest(fit2)

# check the residuals
res2<- residuals(fit2)
plot(res2, xlab='quarter', ylab='residuals', type='l')

acf(res2)
```

```{r}
# data transformed as time series
mac.ts<-ts(imac, frequency=4)

# Model with trend and seasonality
fit3 <- tslm(mac.ts~ trend+season)
summary(fit3)

# check the residuals
res3 <- residuals(fit3)

plot(res3, ylab='residuals')
dwtest(fit3)
```

```{r}
# plot of the model
plot(mac.ts, ylab='iMac sales', xlab='Time')
lines(fitted(fit3), col=2)
```

# Linear regression with trend and seasonality and forecasting exercise

Data on Australian beer production

```{r}
beer<- ausbeer
beer
plot(beer)
Acf(beer)
```

```{r}
#take a portion of data and fit a linear model with tslm
beer1<- window(ausbeer, start=1992, end=2006 -.1)
beer1
plot(beer1)
```

```{r}
m1<- tslm(beer1~ trend+ season)
summary(m1)
fit<- fitted(m1)

plot(beer1)
lines(fitted(m1), col=2)
```

```{r}
fore <- forecast(m1)
plot(fore)
# forecasts from regression model for beer production, The dark shaded region shows 80% prediction intervals and the light shaded 95% prediction intervals (range of values the random variable could take with relatively high probability). 
```

```{r}
# analysis of residuals
res<- residuals(m1) 
plot(res)

# the form of residuals seems to indicate the presence of negative autocorrelation
Acf(res)
```

Data on quarterly percentage change in US consumption, income, production, savings, unemployment

```{r}
uschange<- uschange
str(uschange)
plot(uschange)
autoplot(uschange) 
pairs(uschange)
```

```{r}
#different way of seeing the same series
cons<- uschange[,1]
inc<- uschange[,2]
prod<- uschange[,3]
sav<- uschange[,4]
unem<- uschange[,5]

# consider the series of consumption as dependent variable and study with the other explanatory variables in a multiple regression model
fit.cons<- tslm(cons~inc+prod+sav+unem)
summary(fit.cons)
AIC(fit.cons)

plot(cons)
lines(fitted(fit.cons), col=2)
```

```{r}
res<- residuals(fit.cons)
plot(res)
acf(res)
Acf(res) # note the difference

# we remove the 'production' variable  
fit.cons1<- tslm(cons~inc+sav+unem)
summary(fit.cons1)
```

Exercise with scenario hypotheses

```{r}
# Fit the model again (by using the data in a different way)
fit.consBest <- tslm(
  Consumption ~ Income + Savings + Unemployment,
  data = uschange)

h <- 4  #window for prediction

# we assume a constant increase of 1 and 0.5% for income and savings and no change for unemployment
newdata <- data.frame(
  Income = c(1, 1, 1, 1),
  Savings = c(0.5, 0.5, 0.5, 0.5),
  Unemployment = c(0, 0, 0, 0))

# forecasts
fcast.up <- forecast(fit.consBest, newdata = newdata)

# we assume a constant decrease of 1 and 0.5% for income and savings and no change for unemployment
newdata <- data.frame(
  Income = rep(-1, h),
  Savings = rep(-0.5, h),
  Unemployment = rep(0, h))
fcast.down <- forecast(fit.consBest, newdata = newdata)

# graphical comparison of these two scenarios
autoplot(uschange[, 1]) +
  ylab('% change in US consumption') +
  autolayer(fcast.up, PI = TRUE, series = 'increase') +
  autolayer(fcast.down, PI = TRUE, series = 'decrease') +
  guides(colour = guide_legend(title = 'Scenario'))
```

# Nonlinear models for new product growth (diffusion models)

```{r}
# Music data (RIAA)
music<- read_excel('music.xlsx')
str(music)

# create the variable cassette
cassette<- music$cassette[1:36]

# some simple plots
plot(cassette, type='b')
plot(cumsum(cassette), type='b')
```

```{r}
# a better plot of the yearly time series
plot(cassette, type= 'b',xlab='Year', ylab='Annual sales',  pch=16, lty=3, xaxt='n', cex=0.6)
axis(1, at=c(1,10,19,28,37), labels=music$year[c(1,10,19,28,37)])
```

```{r}
# we estimate a simple Bass Model 
bm_cass<-BM(cassette,display = T)
summary(bm_cass)
```

```{r}
# prediction (out-of-sample)
pred_bmcas<- predict(bm_cass, newx=c(1:50))
pred.instcas<- make.instantaneous(pred_bmcas)

# plot of fitted model
plot(cassette, type= 'b',xlab='Year', ylab='Annual sales',  pch=16, lty=3, xaxt='n', cex=0.6)
axis(1, at=c(1,10,19,28,37), labels=music$year[c(1,10,19,28,37)])
lines(pred.instcas, lwd=2, col=2)
```

```{r}
# we estimate the model with 50% of the data
bm_cass50<-BM(cassette[1:18],display = T)
summary(bm_cass50)
```

```{r}
pred_bmcas50<- predict(bm_cass50, newx=c(1:50))
pred.instcas50<- make.instantaneous(pred_bmcas50)

plot(cassette, type= 'b',xlab='Year', ylab='Annual sales',  pch=16, lty=3, xaxt='n', cex=0.6)
axis(1, at=c(1,10,19,28,37), labels=music$year[c(1,10,19,28,37)])
lines(pred.instcas50, lwd=2, col=2)
```

```{r}
# we estimate the model with 25% of the data
bm_cass75<-BM(cassette[1:9],display = T)
summary(bm_cass75)

pred_bmcas75<- predict(bm_cass75, newx=c(1:50))
pred.instcas75<- make.instantaneous(pred_bmcas75)
```

```{r}
# Comparison between models (instantaneous)
plot(cassette, type= 'b',xlab='Year', ylab='Annual sales',  pch=16, lty=3, xaxt='n', cex=0.6)
axis(1, at=c(1,10,19,28,37), labels=music$year[c(1,10,19,28,37)])
lines(pred.instcas75, lwd=2, col=2)
lines(pred.instcas50, lwd=2, col=3)
lines(pred.instcas, lwd=2, col=4)


# Comparison between models (cumulative)
plot(cumsum(cassette), type= 'b',xlab='Year', ylab='Annual sales',  pch=16, lty=3, xaxt='n', cex=0.6)
axis(1, at=c(1,10,19,28,37), labels=music$year[c(1,10,19,28,37)])
lines(pred_bmcas75, lwd=2, col=2)
lines(pred_bmcas50, lwd=2, col=3)
lines(pred_bmcas, lwd=2, col=4)

###exercise: try the same with the CD time series
```

Twitter (revenues)

```{r}
twitter<- read_excel('twitter.xlsx')
length(twitter$twitter)

plot(twitter$twitter, type= 'b',xlab='Quarter', ylab='Quarterly revenues',  pch=16, lty=3, xaxt='n', cex=0.6)
axis(1, at=c(1,10,19,28,37,46), labels=twitter$quarter[c(1,10,19,28,37,46)])
```

```{r}
###BM
tw<- (twitter$twitter)
Acf(tw)

bm_tw<-BM(tw,display = T)
summary(bm_tw)
```

```{r}
pred_bmtw<- predict(bm_tw, newx=c(1:60))
pred.insttw<- make.instantaneous(pred_bmtw)

plot(cumsum(tw), type= 'b',xlab='Quarter', ylab='Cumulative revenues',  pch=16, lty=3, cex=0.6, xlim=c(1,60), ylim=c(0,40000))
lines(pred_bmtw, lwd=2, col=2)

plot(tw, type= 'b',xlab='Quarter', ylab='Quarterly revenues',  pch=16, lty=3, cex=0.6, xlim=c(1,60))
lines(pred.insttw, lwd=2, col=2)
```

```{r}
# GBMr1
GBMr1tw<- GBM(tw,shock = 'rett',nshock = 1,prelimestimates = c(4.463368e+04, 1.923560e-03, 9.142022e-02, 24,38,-0.1))

# GBMe1
GBMe1tw<- GBM(tw,shock = 'exp',nshock = 1,prelimestimates = c(4.463368e+04, 1.923560e-03, 9.142022e-02, 12,-0.1,0.1))
summary(GBMe1tw)
```

```{r}
pred_GBMe1tw<- predict(GBMe1tw, newx=c(1:60))
pred_GBMe1tw.inst<- make.instantaneous(pred_GBMe1tw)

plot(cumsum(tw), type= 'b',xlab='Quarter', ylab='Cumulative revenues',  pch=16, lty=3, cex=0.6, xlim=c(1,60), ylim=c(0,50000))
lines(pred_GBMe1tw, lwd=2, col=2)

plot(tw, type= 'b',xlab='Quarter', ylab='Quarterly revenues',  pch=16, lty=3, cex=0.6, xlim=c(1,60))
lines(pred_GBMe1tw.inst, lwd=2, col=2)
```
